---
title: "Interactive Employment Transitions with g6r"
author: "Giampaolo Montaletti"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Interactive Employment Transitions with g6r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,  # Set to TRUE when g6R is available
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Introduction to g6r for Employment Transitions

This vignette demonstrates how to create interactive network visualizations of employment transitions using the **g6r** package. The g6r package provides R bindings to the powerful G6 JavaScript visualization library developed by Ant Design, enabling creation of sophisticated interactive network diagrams suitable for exploring complex employment patterns.

## What is g6r?

g6r is an R package that creates interactive network visualizations powered by the G6 graph visualization engine. Key features include:

- **20+ Layout Algorithms**: Force-directed, circular, hierarchical, radial, and more
- **15+ Interactive Behaviors**: Zooming, dragging, selecting, edge creation, brush selection
- **17+ Plugins**: Minimap, tooltips, toolbars, grid lines, edge bundling
- **High Performance**: Handles 20,000+ nodes efficiently
- **Shiny Integration**: Full support for dynamic updates in Shiny applications
- **Accessibility**: Support for colorblind-friendly palettes and high contrast modes

## Installation

```{r install}
# Install from CRAN
install.packages("g6R")

# Or install development version
# pak::pak("cynkra/g6R")

# Also ensure vecshift is loaded
library(vecshift)
library(g6R)
library(data.table)
```

# Basic Interactive Transitions

## Creating Sample Data

Let's start by creating sample employment data and processing it through the vecshift pipeline:

```{r sample-data}
library(data.table)
library(vecshift)

# Create comprehensive sample data
set.seed(42)
employment_data <- data.table(
  id = 1:12,
  cf = rep(c("PERSON001", "PERSON002", "PERSON003"), each = 4),
  INIZIO = as.Date(c(
    "2023-01-01", "2023-04-01", "2023-08-01", "2023-12-01",  # Person 1
    "2023-02-01", "2023-06-01", "2023-09-01", "2023-12-15",  # Person 2  
    "2023-01-15", "2023-05-01", "2023-07-01", "2023-10-01"   # Person 3
  )),
  FINE = as.Date(c(
    "2023-02-28", "2023-06-30", "2023-11-30", "2023-12-31",  # Person 1
    "2023-04-30", "2023-08-31", "2023-11-30", "2023-12-31",  # Person 2
    "2023-03-31", "2023-06-30", "2023-09-30", "2023-12-31"   # Person 3
  )),
  prior = c(1, 0, 1, 1,   # Person 1: Full-time, Part-time, Full-time, Full-time
           1, 1, 0, 1,    # Person 2: Full-time, Full-time, Part-time, Full-time
           0, 1, 1, 0),   # Person 3: Part-time, Full-time, Full-time, Part-time
  company = c(
    "TechCorp", "RetailChain", "Manufacturing", "TechCorp",        # Person 1
    "Healthcare", "TechCorp", "RetailChain", "Manufacturing",      # Person 2
    "RetailChain", "Healthcare", "TechCorp", "RetailChain"         # Person 3
  ),
  salary = c(
    50000, 25000, 55000, 60000,    # Person 1
    65000, 55000, 28000, 58000,    # Person 2
    22000, 70000, 62000, 26000     # Person 3
  ),
  region = c(
    "North", "South", "North", "North",      # Person 1
    "East", "North", "South", "North",       # Person 2  
    "South", "East", "North", "South"        # Person 3
  )
)

print(employment_data)
```

## Processing Through vecshift Pipeline

```{r process-data}
# Process through the vecshift pipeline
pipeline_result <- process_employment_pipeline(
  original_data = employment_data,
  merge_columns = c("company", "salary", "region"),
  show_progress = TRUE
)

# Check the processed data
print(head(pipeline_result, 10))
```

## Analyzing Employment Transitions  

```{r analyze-transitions}
# Analyze company transitions with salary and region statistics
company_transitions <- analyze_employment_transitions(
  pipeline_result = pipeline_result,
  transition_variable = "company",
  statistics_variables = c("salary", "region"),
  min_unemployment_duration = 1,
  show_progress = TRUE
)

print(company_transitions)

# Also create a transition matrix for comparison
transition_matrix <- analyze_employment_transitions(
  pipeline_result = pipeline_result,
  transition_variable = "company",
  output_transition_matrix = TRUE
)

print(transition_matrix)
```

## Basic Interactive Visualization

```{r basic-g6r}
library(g6R)

# Create basic interactive visualization using package function
basic_plot <- plot_interactive_transitions(
  transitions_data = company_transitions,
  layout = "force",
  height = "600px"
)

basic_plot
```

## Exploring Different Layouts

```{r layouts}
# Force-directed layout (good for organic network exploration)
force_plot <- plot_interactive_transitions(
  company_transitions,
  layout = "force",
  show_minimap = TRUE,
  height = "500px"
)

# Circular layout (good for comparing relationships)
circular_plot <- plot_interactive_transitions(
  company_transitions,
  layout = "circular", 
  show_minimap = FALSE,
  height = "500px"
)

# Hierarchical layout (good for directed flows)
hierarchical_plot <- plot_interactive_transitions(
  company_transitions,
  layout = "dagre",
  height = "500px"
)

# Display the plots
force_plot
circular_plot  
hierarchical_plot
```

# Advanced Customization

## Customizing Node and Edge Appearance

```{r customization}
# Customize based on different metrics
custom_plot <- plot_interactive_transitions(
  transitions_data = company_transitions,
  layout = "force",
  # Use transition duration for edge width instead of count
  edge_width_metric = "transition_duration",
  # Use incoming transitions for node size
  node_size_metric = "in_degree",
  # Filter to significant transitions only
  min_weight_threshold = 2,
  # Enable edge bundling for cleaner visualization
  edge_bundling = TRUE,
  # Custom colors
  node_color_palette = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd"),
  edge_color = "#999999"
)

custom_plot
```

## Working with Transition Matrices

```{r matrix-format}
# Convert matrix format to interactive visualization
matrix_plot <- plot_interactive_transitions(
  transitions_data = transition_matrix,
  layout = "circular",
  show_labels = TRUE,
  enable_select = TRUE
)

matrix_plot
```

## Accessibility Features

```{r accessibility}
# Create colorblind-friendly visualization
accessible_plot <- plot_interactive_transitions(
  transitions_data = company_transitions,
  layout = "force",
  accessibility_mode = TRUE,
  # High contrast colorblind-friendly palette
  node_color_palette = c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                        "#CC79A7", "#F0E442", "#0072B2", "#D55E00"),
  show_labels = TRUE,
  height = "600px"
)

accessible_plot

# Test different accessibility scenarios
accessibility_tests <- test_g6r_accessibility(
  transitions_data = company_transitions,
  test_scenarios = c("colorblind", "high_contrast", "large_nodes")
)

# View each accessibility test
accessibility_tests$colorblind_safe
accessibility_tests$high_contrast
accessibility_tests$large_elements
```

# Shiny Integration

## Basic Shiny Application

```{r shiny-basic}
library(shiny)
library(g6R)

# Create a basic Shiny app for exploring transitions
ui <- fluidPage(
  titlePanel("Interactive Employment Transitions Explorer"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("layout", "Choose Layout:",
                 choices = list(
                   "Force-directed" = "force",
                   "Circular" = "circular",
                   "Hierarchical" = "dagre",
                   "Radial" = "radial"
                 ),
                 selected = "force"),
      
      numericInput("min_weight", "Minimum Transition Weight:",
                  value = 1, min = 1, step = 1),
      
      checkboxInput("show_minimap", "Show Minimap", TRUE),
      checkboxInput("accessibility_mode", "Accessibility Mode", FALSE),
      
      hr(),
      
      h4("Export Options"),
      downloadButton("export_data", "Download Data", class = "btn-primary")
    ),
    
    mainPanel(
      g6Output("transition_plot", height = "600px"),
      
      br(),
      
      h4("Transition Statistics"),
      tableOutput("stats_table")
    )
  )
)

server <- function(input, output, session) {
  
  # Reactive data processing
  transitions_data <- reactive({
    analyze_employment_transitions(
      pipeline_result = pipeline_result,
      transition_variable = "company",
      statistics_variables = c("salary", "region"),
      min_unemployment_duration = input$min_weight
    )
  })
  
  # Main visualization
  output$transition_plot <- renderG6({
    plot_interactive_transitions(
      transitions_data = transitions_data(),
      layout = input$layout,
      show_minimap = input$show_minimap,
      accessibility_mode = input$accessibility_mode,
      min_weight_threshold = input$min_weight
    )
  })
  
  # Statistics table
  output$stats_table <- renderTable({
    data <- transitions_data()
    
    if (nrow(data) > 0) {
      data.frame(
        Transition = paste(data$from, "→", data$to),
        Count = data$weight,
        `Avg Unemployment (days)` = round(data$transition_duration, 1),
        check.names = FALSE
      )
    }
  })
  
  # Download handler
  output$export_data <- downloadHandler(
    filename = function() {
      paste0("transitions_", Sys.Date(), ".csv")
    },
    content = function(file) {
      write.csv(transitions_data(), file, row.names = FALSE)
    }
  )
}

# Run the app (uncomment to run)
# shinyApp(ui, server)
```

## Advanced Shiny Module

The package includes a complete Shiny module for employment transitions exploration:

```{r shiny-module}
# Use the built-in module
library(shiny)

ui <- fluidPage(
  titlePanel("Advanced Employment Transitions Explorer"),
  
  # Use the pre-built module
  interactive_transitions_module()$ui("transitions")
)

server <- function(input, output, session) {
  
  # Your transition data (reactive)
  transition_data <- reactive({
    analyze_employment_transitions(
      pipeline_result = pipeline_result,
      transition_variable = "company",
      statistics_variables = c("salary", "region")
    )
  })
  
  # Call the module server
  interactive_transitions_module()$server("transitions", transition_data)
}

# Run the advanced app (uncomment to run)
# shinyApp(ui, server)
```

# Advanced Use Cases

## Comparing Groups

```{r group-comparison}
# Create data for different regions
north_data <- employment_data[region == "North"]
south_data <- employment_data[region == "South"]

# Process each region separately
north_result <- process_employment_pipeline(north_data, merge_columns = "company")
south_result <- process_employment_pipeline(south_data, merge_columns = "company")

# Analyze transitions for each region
north_transitions <- analyze_employment_transitions(
  north_result, 
  transition_variable = "company",
  show_progress = FALSE
)

south_transitions <- analyze_employment_transitions(
  south_result,
  transition_variable = "company", 
  show_progress = FALSE
)

# Create comparison visualizations
regional_comparison <- compare_transitions_between_groups(
  transitions_list = list(
    "North Region" = north_transitions,
    "South Region" = south_transitions
  ),
  layout = "circular",
  min_weight_threshold = 1
)

# Display each region's transitions
regional_comparison$`North Region`
regional_comparison$`South Region`
```

## Time-Based Analysis

```{r time-analysis}
# Simulate data for different years
data_2022 <- employment_data[year(INIZIO) == 2023]  # Using 2023 as proxy for demo
data_2022[, `:=`(INIZIO = INIZIO - 365, FINE = FINE - 365)]

# Process each year
result_2022 <- process_employment_pipeline(data_2022, merge_columns = "company")
result_2023 <- process_employment_pipeline(employment_data, merge_columns = "company")

# Create time-based visualization (foundation for animation)
time_based_plot <- plot_transitions_over_time(
  pipeline_results = list(result_2022, result_2023),
  time_periods = c("2022", "2023"),
  transition_variable = "company",
  animation_speed = 2000
)

time_based_plot
```

## Large-Scale Data Performance

```{r large-scale}
# Generate larger dataset for performance testing
large_demo_data <- generate_g6r_demo_data(
  n_persons = 200,
  n_companies = 20,
  time_span_years = 2,
  transition_probability = 0.4
)

print(paste("Generated", nrow(large_demo_data), "employment records"))

# Process large dataset
large_pipeline_result <- process_employment_pipeline(
  large_demo_data,
  merge_columns = c("company", "region", "employment_type")
)

# Analyze transitions (filter to significant ones)
large_transitions <- analyze_employment_transitions(
  large_pipeline_result,
  transition_variable = "company",
  statistics_variables = c("salary", "region"),
  min_unemployment_duration = 7  # Focus on longer unemployment gaps
)

print(paste("Found", nrow(large_transitions), "unique transition patterns"))

# Create visualization optimized for performance
large_scale_plot <- plot_interactive_transitions(
  transitions_data = large_transitions,
  layout = "force",
  min_weight_threshold = 3,  # Filter to significant transitions
  edge_bundling = TRUE,      # Reduce visual clutter
  show_minimap = TRUE,       # Help navigation
  height = "700px"
)

large_scale_plot
```

# Data Export and Integration

## Exporting g6r Data

```{r export}
# Convert transitions to g6r format for inspection
g6_data <- convert_transitions_to_g6r(company_transitions)

# Examine the node structure
print("Node data:")
print(head(g6_data$nodes))

print("Edge data:")
print(head(g6_data$edges))

# Export for external use
write.csv(g6_data$nodes, "employment_transitions_nodes.csv", row.names = FALSE)
write.csv(g6_data$edges, "employment_transitions_edges.csv", row.names = FALSE)
```

## Integration with Other R Packages

```{r integration}
# Convert g6r data for use with other network packages
library(igraph, warn.conflicts = FALSE)

# Create igraph object from g6r data  
graph_data <- g6_data
edges_igraph <- graph_data$edges[, c("source", "target", "weight")]

# Create igraph network
ig <- graph_from_data_frame(
  d = edges_igraph, 
  vertices = graph_data$nodes,
  directed = TRUE
)

# Calculate network metrics
print("Network metrics:")
print(paste("Nodes:", vcount(ig)))
print(paste("Edges:", ecount(ig)))
print(paste("Density:", round(edge_density(ig), 3)))
print(paste("Components:", components(ig)$no))

# Calculate centrality measures that can inform g6r visualization
centrality_measures <- data.frame(
  node = V(ig)$name,
  betweenness = betweenness(ig, normalized = TRUE),
  closeness = closeness(ig, normalized = TRUE),
  eigenvector = eigen_centrality(ig)$vector
)

print("Top nodes by betweenness centrality:")
print(head(centrality_measures[order(-centrality_measures$betweenness), ]))
```

# Best Practices and Performance Tips

## Performance Optimization

1. **Filter Data Appropriately**: Use `min_weight_threshold` to focus on significant transitions
2. **Choose Layouts Wisely**: Force-directed layouts are computationally intensive for large networks
3. **Enable Edge Bundling**: Reduces visual clutter and improves rendering performance
4. **Limit Simultaneous Features**: Don't enable all plugins at once for large datasets

## Accessibility Guidelines

1. **Use Colorblind-Friendly Palettes**: Always test with simulated color vision deficiency
2. **Provide Redundant Encoding**: Use size + color, or shape + color for critical information
3. **Ensure Sufficient Contrast**: Test with WCAG AA standards (4.5:1 minimum)
4. **Include Alternative Text**: Document what the visualization shows for screen readers

## Shiny Performance

1. **Use Reactive Efficiently**: Cache expensive computations
2. **Implement Progressive Loading**: Start with filtered data, allow expansion
3. **Debounce User Inputs**: Prevent excessive re-rendering during slider adjustments
4. **Provide Export Options**: Allow users to save visualizations and data

# Troubleshooting

## Common Issues

### g6R Package Installation
```{r troubleshooting-install}
# If installation fails, try:
# install.packages("g6R", dependencies = TRUE)
# 
# For development version:
# remotes::install_github("cynkra/g6R")
#
# Check installation
if (requireNamespace("g6R", quietly = TRUE)) {
  message("g6R is available")
} else {
  message("g6R is not installed - install with: install.packages('g6R')")
}
```

### Empty Visualizations
```{r troubleshooting-empty}
# Check your data has transitions
check_transitions <- function(transitions_data) {
  if (is.matrix(transitions_data)) {
    total_transitions <- sum(transitions_data)
    message("Matrix format: ", total_transitions, " total transitions")
  } else if (is.data.frame(transitions_data) || data.table::is.data.table(transitions_data)) {
    message("Data.table format: ", nrow(transitions_data), " transition patterns")
    message("Total transitions: ", sum(transitions_data$weight, na.rm = TRUE))
  }
  
  if (nrow(transitions_data) == 0 || (is.matrix(transitions_data) && sum(transitions_data) == 0)) {
    message("No transitions found. Check your min_unemployment_duration parameter.")
  }
}

check_transitions(company_transitions)
```

### Performance Issues
```{r troubleshooting-performance}
# For large datasets, optimize by:

# 1. Pre-filtering data
optimized_transitions <- company_transitions[weight >= 2]  # Focus on frequent transitions

# 2. Using simpler layouts
simple_plot <- plot_interactive_transitions(
  transitions_data = optimized_transitions,
  layout = "circular",  # Faster than force-directed
  show_minimap = FALSE, # Reduces rendering load
  edge_bundling = FALSE # Disable for better performance
)
```

# Conclusion

The g6r package provides powerful capabilities for creating interactive employment transition visualizations. Key advantages include:

- **Rich Interactivity**: Users can explore data through zooming, selecting, and filtering
- **Professional Appearance**: High-quality visualizations suitable for presentations and reports
- **Shiny Integration**: Seamless embedding in interactive web applications
- **Accessibility Support**: Built-in features for inclusive design
- **Performance**: Efficient rendering of large networks

This integration with the vecshift package creates a complete pipeline from raw employment data to interactive visualizations, enabling deep exploration of career transition patterns and employment dynamics.

## Next Steps

- Explore custom JavaScript integration for advanced features
- Implement real-time data updates in Shiny applications
- Create dashboard templates for common employment analysis scenarios
- Develop time-series animation capabilities
- Add statistical overlays and clustering visualizations

For additional examples and advanced features, see the package documentation and the g6R project website at https://cynkra.github.io/g6R/.